Museum specimens as baseline for assessing a potential global pollinator crisis

### 2018-03-08 ms.Rmd
added a section on the NZ study and have edited the introduction and abstract

There is an ever-increasing concern about the current decline of pollinators, specifically bees. Despite pollinator decline reports are widespread, the information available is scarce and biased geographically and taxonomically, limiting the inference about a potential pollination crisis. The most direct form for understanding species decline patterns is to analyze historical time series, however, those are rare or often non-existing for most pollinator groups. However, museum collections can provide the only source of historical data serving as a baseline to understand the reach of pollinator declines. Specimens historically collected by naturalists and preserved in museums can provide information on when and where species were collected, as well as other ecological information like historical species interactions, or morphological traits. We show here how researchers have used historical museum specimens to compare historical occurrences with current occurrence patterns to show changes over time in biodiversity levels, species relative abundances, body size and even pollination services delivery. Despite this advances, we show that we still miss information for the status and trends of most pollinators and that often a bottle neck is digitizing the historical data. We analyze the opportunities and limitations to do so worldwide and provide two contrasting worked examples on how to analyze this type of data for two distinct regions for which historical bee declines have never been assessed, New Zealand and Spain. There is a great potential for Museum specimens to play a key role in elucidating the extent of the pollinator crisis and compare the responses obtained along different environmental situations if we succeed in unleashing the power of historical data.

@@

Pollinators are of paramount importance given its prominent role in plant reproduction (@ollerton2012) and food security (@kleinetal2007). As many other taxa, pollinators are also exposed to a plethora of human induced rapid environmental changes potentially threatening its populations. This conundrum was realized early on by researchers, and was popularized by the influential book "the forgotten pollinators" (1996). However, at that time there was almost no data on the trends and status of pollinators and this initial claims trigger the first efforts to elucidate the problem, including a commission of a US National Academy of Science (NAS) panel to assess the reach of pollinators declines (2006). The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, the studies on pollinator responses to global change has multiplied and we reached a strong consensus that pressures like habitat destruction, land use change, chemical exposure, exotic species and global warming are the leading causes of pollinator declines and act sinergistically (@goulson2016, @potts2010). However, the status and trends of most pollinator species remains unknown. For example, for a well studied region like Europe, 60% [check number]() of the bee species assessed by a recent IUCN report fall into the category "Data Deficient" (@Nieto2014).

One of the main barriers to profile pollinator population trends is that pollinators conform an heterogeneous group including insects such as bees, flies, butterflies and beetles, but also birds, bats and even lizards (@Winfreeetal2011). In particular bees, usually regarded as the more important in terms of abundance, efficiency and widespread distribution (@ollerton2018), are a diverse group with more than 20 000 species worldwide (@PickeringAscher) and most of them require expert taxonomic expertise for its identification. This has led not only to a biased geographic patterns due to the uneven distribution of researchers (@Martinetal2012), but also a taxonomic bias to species easier to identify, like bumblebees (@ArcherOikos).

The second barrier is methodological and involves how to assess population trends for highly mobile organisms, which in addition are often short-lived and small. A practical approach has been to use space for time substitutions, where the researchers compares natural areas with areas affected by different environmental conditions. Despite the critiques on the robustness of this approach (@ref), this is the larger source of information we have, specially for Europe, where recent joint efforts had estimated bee richness declines at the country level using the predictions from such models (@DePalma). A second promising method is pollinator monitoring programs. Inspired in the successful butterfly monitoring programs (@ref), this protocols, often carried by citizen scientists, are being implemented on other taxa (@ref). The main limitation is that those programs need time to build up long term data series and do not reveal anything about the original conditions. Finally, the best source of historical data can be found in museums. The core aim of Museums is precisely to conserve and curate this kind of collections, and serve not only of a precious voucher of specimens, but often also assures its taxonomic identification quality (@primak). The major bottleneck to use this information, is however, digitizing the specimens. Efforts to do so have provide great benefits and allow the characterization of the status and trends of bee pollinators is the US and EU (e.g. Bartomeus et al 2013, Scheper et al 2014) and despite data deposited in museums is often collected opportunistically and have its own biases (@Keil), is the only data we have to reconstruct the past.

Regardless of the method, it's important to consider that biodiversity has many aspects beyond species richness, and we should put more effort in differentiating declines not only in local (alpha) diversity or regional (gamma) diversity, but also in the turnover over sites (beta), the abundance of key species, and the evenness of the communities (@winfree2016ELE). In fact, environmental change often give us few winners, and lots of loosers (@Bartomeusetal2013), and characterizing those is extremely important as the few winners are often exotic species (@Stavertetal2018).  [Add something about traits of winners and loosers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296
[move this and expand to the discussion on the map?]()

@@

Digitizing old specimens is not a trivial task and it needs expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geolocate the coordinates of the collection event (@ref and software) and (iii) store the data in a proper curated database (@ref on database management). Doing this process one by one with each of the thousands of specimens located in a museum is a daunting task and needs specialized personnel. While some tasks can only be done by expert taxonomists, new technologies and citizen science can speed up the process of digitizing old collections. High resolution photos and scanning of the associated labels have been uploaded to the internet to get the power of distributed work where thousands of volunteers can transcript the images to text. New algorithms have emerged to georeference localities based on vernacular names (@refs geonames, geoparser and opencage). Explain zooniverse, NZ initiative and how data can be validated... [Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]
[how much do we want to extend on this? Is interesting info, but boring o read?]()

@@

Is not surprising the the best evidence of pollinator declines comes from the more disturbed countries. A pioneer study using citizen science and museum specimens in UK and Netherlands, two of the most populated and transformed countries showed a worrisome trend for both bees, flies and flowering plants, but specially for bees (@Beisjmieretal2006). However, it's important to note that even within those countries, the local estimates are all nearby big cities and agricultural areas, and data is lacking for the better conserved areas. Further exploration of this data set shows that the declining trend has attenuated in recent decades (@carvalheiroetal). Detailed trends for common bee species in the Netherlands using museum specimens confirm the link between pollinator and plant declines, as bees wich preferred host plants have declined the most are the more threatened (@Schepper2014).   
[- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies! Add Ollerton paper!]()

In comparison, regional estimates for the eastern US using museum specimens show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions as most species can still persist (@Colla2012?). This is in harsh contrast with the high local extinctions reported by Burkle (@Burkle) when comparing historical observations of a large forested ecosystem with the few remnants that still exists, but note that this comparison do not takes into account the opportunistic bees using other land use changes, and that the local richness may have changes, but not declined as drastically. In any case, there is a great concordance between species locally extinct and regionally declining (@Bartomeus_winfree2013) indicating that local extinctions are part of the downward regional population trends.

Outside of Europe and the US there is surprisingly little evidence of pollinator declines using historical records. [South Africa: Pauw, maybe with plants...]() .  However, there are specific examples showing the historical decline of certain species e.g. Bombus dalbhomi (Aizen)

@@

Justify focus on bees? We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species declines (b), syrphid species declines (c) and ?? to highlight the little coverage we have...

@@

## Case study two: NZ

- Describe data and analysis.   
- Present main results

@@

E. Asensio (or maybe coauthor?). Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL.

>>>>

There is ever-increasing concern about the decline of pollinators, especially for bees. However, despite reports that pollinator declines are widespread, data is scarce and geographically and taxonomically biased. This limits any robust inference about a potential pollination crisis. The most direct approach for identifying patterns in species declines is analysis of historical time series data. However, these data are often rare or non-existent for most pollinator groups, but provide the only source of historical information serving as a baseline to understand the extent of pollinator declines. Specimens historically collected by naturalists and preserved in museums can provide information on when and where species were collected, as well as other ecological information (e.g., historical species interactions and morphological traits). Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current occurrence patterns to identify changes in biodiversity, relative species abundances, body size and even pollination services over time. Despite recent advances, we show that information on the status and trends of most pollinators is missing and a key bottle neck slowing progress in this field is the digitization of historical data. We identify opportunities and limitations to progress understanding of pollinator declines worldwide. Further, we demonstrate how to analyse historical museum collection data, by providing two contrasting case studies from distinct geographical regions, New Zealand and Spain, for which historical pollinator declines have never been assessed. We show that there is a great potential for museum specimens to play a key role in identifying the extent of the pollination crisis if focus on unleashing the power of historical collection data.

@@

Animal pollinators are acritical component of ecosystems, given their role in plant reproduction (@ollerton2012) and food security worldwide (@kleinetal2007). As many other taxa, pollinators are vulnerable to multiple human-mediated environmental changes, which potentially threatening their populations. This problem was identified by researchers several decades ago, and was popularized by the influential book "the forgotten pollinators" (1996). However, these accounts were relatively anecdotal as there were almost no data on pollinator trends and statuses at the time. These initial claims triggered the first efforts to address the problem, including a US National Academy of Science (NAS) panel commissioned to assess the extent of pollinators declines (2006). The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, studies on pollinator responses to global change have multiplied and researchers have reached strong consensus that pressures such as habitat destruction, land use change, chemical exposure, exotic species and climate change are the primary causes of pollinator declines, and often act synergistically (@goulson2016, @potts2010). Yet the status and trends of most pollinator species worldwide remain unknown. For example, even for a well-studied region such as Europe, 60% [check number]() of bee species assessed by in recent IUCN report fall into the category "Data Deficient" (@Nieto2014). For countries outside of Europe and the US, data on pollinator population is almost non-existent.

One of the main barriers to identifying pollinator population trends is that pollinators are a diverse group of taxa including bees, flies, butterflies and beetles, and also birds, bats and even lizards (@Winfreeetal2011). In particular, bees are usually regarded as the more important pollinaotrs due to their abundance, pollination efficiency and widespread distribution (@ollerton2018). However, bees are an incredibly diverse group, with more than 20,000 species worldwide (@PickeringAscher), and most require taxonomic expertise for identification. This has led to a biased geographic patterns in bee decline research due to the uneven distribution of researchers (@Martinetal2012), but also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos). Another barrier is methodological and involves challenges for assessing population trends for highly mobile organisms, which are also often short-lived and small.
One solution to overcoming these barriers has been to use space for time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data currently available and in Europe, researchers have recently estimated bee richness declines for each country using predictions generated from models (@DePalma). A second promising method is pollinator monitoring programs, with data often collected by citizen scientists. This approach was inspired by successful butterfly monitoring programs and is currently being extended to other pollinator taxa (@ref). The main limitation is that these programs need significant time to generate long-term datasets and do not reveal anything about historic pollinator communities. Finally, the most robust approach for identifying historical pollinator tends is to use museum collection data. The core aim of museums is precisely to conserve and curate collections, thus they not only serve as a precious repository specimens, but often also ensure high quality taxonomic identification (@primak). However, the major bottleneck for researchers wanting to use this information is the digitization of specimens. Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have facilitated identification of current status and trends in pollinator populations (e.g. Bartomeus et al 2013, Scheper et al 2014). This is despite the fact that museum collections often have a number of biases (@Keil). In reality, museum collections are the only data we have to reconstruct the past.

Regardless of the method, it is important to consider that biodiversity has many aspects beyond species richness, and we should into identifying changes not only in local (alpha) diversity or regional (gamma) diversity, but also in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE). Environmental changes often results in few winners and many of losers (@Bartomeusetal2013). Identifying winners and losers is critical as the few winners are often exotic species, and represent a subset of trait that facilitate survival in highly modified environments (@Stavertetal2018). These changes can have important flow on effects for pollination of native species and crops.  [Add something about traits of winners and losers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296
[move this and expand to the discussion on the map?]()

@@

Digitizing old specimens is not a trivial task and it needs expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of the collection event (@ref and software) and (iii) store the data in a properly curated database (@ref on database management). Undertaking this process for the thousands of specimens a museum collection is a daunting task and requires specialized personnel. While some tasks can only be undertaken by those with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the process of digitizing collections. High resolution photos of specimens and the associated labels can be uploaded to the internet, where the work of transcribing the images can be distributed across hundreds or thousands of volunteers. Furthermore, new algorithms have been created to geo-reference localities based on vernacular names (@refs geonames, geoparser and opencage). Explain zooniverse, NZ initiative and how data can be validated... [Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]
[how much do we want to extend on this? Is interesting info, but boring o read?]()

@@

Is not surprising the best evidence of pollinator declines comes from the more disturbed countries western countries. For the UK and the Netherlands, which have been subject to extensive environmental changes, a study using citizen science and museum specimens found declines for bees, flies and flowering plants (@Beisjmieretal2006). However, it is important to note that even for these two countries, the local estimates for pollinator richness are nearby all for large cities and agricultural areas, and data is lacking for conserved natural areas. Further exploration of this dataset showed that the declining trend has attenuated in recent decades (@carvalheiroetal). Using museum specimens in the Netherlands, detailed trends for common bee species confirm the link between pollinator and plant declines, as bees which preferred host plants have declined the most are the more threatened (@Schepper2014).??confusing.    
[- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies! Add Ollerton paper!]()

In comparison, regional estimates in species richness changes for the eastern US show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions as most species tend to persist (@Colla2012?). These findings are in stark contrast with the widespread local extinctions reported by Burkle (@Burkle), who compared historical observations of bees from a large forested ecosystem with the few remnants. However, it is important to note that this comparison does not take into account opportunistic bee species using novel land use types in the region. Thus, local richness may have declines but regional richness may not have declined as drastically. In any case, there is strong concordance between local extinctions and regional declines (@Bartomeus_winfree2013), suggesting that local extinctions are indicators of regional population declines.

Outside of Europe and the US there is surprisingly studies on pollinator declines using historical records. [South Africa: Pauw, maybe with plants...]() .  However, there are specific examples showing the historical decline of certain species e.g. Bombus dalbhomi (Aizen)

@@

Justify focus on bees? We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes to highlight the little coverage that we have globally.

@@

##Case study two: NZ    
New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation. Much of New Zealand’s pollinator fauna is also relatively depauperate. For example, New Zealand has only 27 species of native bees (@Donovan, 2007), which is a fraction of nearby Australia’s c. 1600 species (REF). However, New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems (@Newstrom & Robertson, 2005). Therefore, New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.

Human colonisation of New Zealand was relatively recent in global terms (c. 740 y) (@Wilmshurst et al. 2008). Before human arrival, New Zealand was mostly forested, but has since been dramatically altered by people. Early Māori settlers cleared forests by burning and more recently, European colonists cleared much of remaining forests for agriculture, mostly before 1900 (@Perry et al, 2014). Therefore, human activity likely affected New Zealand pollinator communities long before collections began. Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.

New Zealand pollinator collections
We used New Zealand bee collection records gathered from multiple sources, including university, research institute, museum and private collections. Fly pollinator data was obtained from six participating New Zealand museums and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators. Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).

Data analyses
First, we filtered our original datasets so that data used for analyses only included unique collection events. To do this, we removed specimens collected at the same location, on the same data and by the same collector. We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980. We accounted for differences in collection effort through time by binning collection records, so that each bin had a similar number of records but a different number of years (see @Bartomeusetal2013 for further details). We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin. Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness. Thus reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.

To test whether there were relative differences in occurrences of species over time (should we analyse natives and exotics together or separately?). To determine if the probability of finding a species in the collection changed over time we used a general linear model with a binomial distribution and a logit link. For species that showed overdispersion, we used a quasi-binomial distribution. We only included species in this analysis that had more than 30 records. To account for differences in sampling effort between years, we weighted each year by the total number of samples collected in that year.

Changes in species richness over time 
We found that rarefied richness for both native and exotic bees increased overtime, but these trends were non-significant (P-values for both groups > 0.05). In contrast, native fly richness declined, whereas exotic fly richness increased, although results for these groups were also non-significant (P-values for both groups > 0.05).

Changes in species relative abundances over time 
We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic). Three bee species declined in relative abundance (one native and two exotic). However, only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic).

Fig 3: Multi-panel figure

-	Do we need some sort of separate discussion or integrate into the results section?
… Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness.

@@

E. Asensio (or maybe coauthor?). Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL. Barry Donovan for providing New Zealand bee collection records. Carola Warner for entering the New Zealand data.
